---
title: "project1"
format: html
---

```{r}
library(tidyverse)
library(Distance)

d <- read_csv("A1_data.csv")
```

# Notes

- truncation distance is 200m
- distances are recorded in 10m categories 
    (5m recoreded distance ~ in the first 10m band)
- randomly selected 1km square plots, about 1% of total plots used
- 2 x 1km line transects in each plot


# Wrangling

```{r}
d <- d |>
  mutate(distance = distance_cat,
         area = 78000,
         region = "A")
```

Sample.Label     # ID number of the 1km square that was surveyed
object_id        # ID number of the observation for each individual bird detected
distance_cat     # distance in m of the bird from the transect line. this is the midpoint of the 10m wide distance bands. 
Urban            # the percentage of the 1km square that is Urban habitat (min = 0, max = 100)
Coniferous       # the percentage of the 1km square that is Coniferous forest (min = 0, max = 100)
Broadleaf        # the percentage of the 1km square that is Broadleaf forest (min = 0, max = 100)
MountainHeathBog # the percentage of the 1km square that is Mountain, Heath, or Bog (min = 0, max = 100)
Freshwater       # the percentage of the 1km square that is Freshwater habitat (min = 0, max = 100)

Q1. Please use the package Distance to produce your best distance sampling model of the data for each species. Demonstrate that it is a good model and use it to estimate the abundance of each of the the two species in Scotland and associated confidence intervals.

```{r}
d_a <- d |>
  filter(species == "A",
         !is.na(distance_cat))

d_b <- d |>
  filter(species == "B",
         !is.na(distance_cat))
```

```{r}
region_table <- tibble(Region.Label = "A",
                       Area = 78000)

sample_table_a <- unique(tibble(Sample.Label = d_a$Sample.Label, 
                         Region.Label = d_a$region,
                         Effort = 2))

obs_table_a <- tibble(object = d_a$object_id,
                      Region.Label = d_a$region,
                      Sample.Label = d_a$Sample.Label)

sample_table_b <- unique(tibble(Sample.Label = d_b$Sample.Label, 
                         Region.Label = d_b$region,
                         Effort = 2))

obs_table_b <- tibble(object = d_b$object_id,
                      Region.Label = d_b$region,
                      Sample.Label = d_b$Sample.Label)
```


# EDA

```{r}
# ggplot(d_a, aes(x = distance_cat)) +
#   geom_histogram(bins = 8) +
#   theme_bw()
```

# Modeling

```{r}
a_hr_cos <- ds(
  data = d_a,
  truncation = 200,
  transect = "line",
  formula = ~ Urban + Coniferous + Broadleaf + MountainHeathBog + Freshwater,
  cutpoints = seq(0, 200, 10),
  key = "hn",
  # adjustment =  "cos",
  region_table = region_table,
  sample_table = sample_table_a,
  obs_table = obs_table_a,
  convert_units = 0.001)

gof_ds(a_hr_cos)

plot(a_hr_cos)
```
```{r}
dht(a_hr_cos$ddf, 
    obs.table = obs_table_a,
    region.table = region_table,
    sample.table = sample_table_a)
```

```{r}
b_hr_cos <- ds(
  data = d_b,
  truncation = 200,
  transect = "line",
  formula = ~ Urban + Coniferous + Broadleaf + MountainHeathBog + Freshwater,
  cutpoints = seq(0, 200, 10),
  key = "hn",
  # adjustment =  "cos",
  region_table = region_table,
  sample_table = sample_table_b,
  obs_table = obs_table_b,
  convert_units = 0.001)

gof_ds(b_hr_cos)

plot(b_hr_cos)
```

```{r}
dht(b_hr_cos$ddf, 
    obs.table = obs_table_b,
    region.table = region_table,
    sample.table = sample_table_b)
```

# Questions

## 1

Please use the package Distance to produce your best distance sampling model of the data for each species. Demonstrate that it is a good model and use it to estimate the abundance of each of the the two species in Scotland and associated confidence intervals.

## 2

It took a lot of effort to collate the habitat data and we’re considering not collecting it in future. Can you provide advice about whether you recommend we continue to collect habitat data? (You can use the results from models)

## 3

There is evidence from other studies that some birds are louder in Urban environments, but there is also more background noise. Can you describe how we might test the effects of these two factors on detectability? Also explain and justify whether this investigation is important for accurately estimating abundance.

Volume differences from the bird calls and from the background noise in urban areas compared to other study areas is a potentially important consideration for abundance estimation. However, the raised background noise and bird volume does not directly violate any assumptions of distance sampling: perfect detection at a single distance (typically 0 meters) should not be affected, as birds on the line should still be just as detectable in slightly noisier conditions. The difference in background noise specifically should not affect the uniform distribution of animals or placement of lines. Independence and error of observations should also be relatively unaffected, but it is possible that animals are more likely to be closer together to communicate through the noise. That would need to be ruled out for strict adherence to the assumptions. The error associated with visual detection would be unaffected, but if auditory detection is necessary, lack of error in the urban areas would need to be proved. Non-reaction to human presence should be simple to justify, as the urban population is more used to human activity. Assuming those minor caveats can be dismissed, then since we still meet all assumptions of distance sampling, the potential drop in detectability in urban areas would be irrelevant given that a habitat covariate was included to account for that variability.

## 4

I’m trying to learn about distance sampling and it seems like there are a lot of assumptions which are not always met. Can you briefly describe a paper that develops an adaptation to the basic distance sampling model to address violations of a distance sampling assumption? Describe the assumption, the implications of violating the assumption, and a brief description of the method developed. Find a paper that we have not explored in lectures.

The 2025 article by Duane R. Diefenbach et al. titled "Accounting for non-random samples with distance sampling to estimate population density" describes a method for overcoming the violation of a core assumption of distance sampling: distance measurements are uniform random variables. This assumption states that lines are equally likely to be placed at any distance from the animal, and animals are equally likely to be at any distance from the line. This is a crucial assumption, as without it, we can not separate the change in detection probability due to distance, from the change in detection probability due to difference in abundance of animals. With those two processes contributing to the detections, we can not rule out bias in either direction.

Violation of this assumption is a relevant problem to solve, as many studies are conducted in areas in which human or natural landscape impediments exist, which may increase or decrease animal abundance in those areas. Those impediments may also dictate the placement of transect lines. 
  
This paper presents a method for correcting this bias using auxiliary data describing a density gradient and a separate resource selection model to estimate the population proportion in the survey area, which is used to estimate abundance in the total area. Estimation of the density gradient may come from GPS locations of sighted animals. The resource selection model may take the form of a mixed-effects logistic regression model with covariates describing the distance from landscape features, habitat type, etc.

This was a very brief overview of the methods presented in Diefencbach et al. 2025, the methods section, and the White-Tailed Deer case study include lots of information for the interested reader.

Diefenbach, D. R., Trowbridge, J., Van Buskirk, A., McConnell, T., Lamp, K., Marques, T. A., Walter, W. D., Wallingford, B. D., & Rosenberry, C. S. (2025). Accounting for non-random samples with distance sampling to estimate population density. Journal of Applied Ecology, 62, 986–994. https://doi.org/10.1111/1365-2664.70006 



























