---
title: "A3_Report"
format: html
---

# Packages

```{r}
library(tidyverse)
library(unmarked)
library(sf)
library(maps)
```

Code from the scotmaps R project. https://github.com/fozy81/scotmaps/blob/master/vignettes/scotmaps.Rmd 

```{r}
install.packages("devtools")
devtools::install_github("fozy81/scotmaps")
library(scotmaps)
ma <- get_layer("marine_areas")
```


# Data

```{r, message=FALSE}
tot_landcover <- read_csv("landcover_total_scotland.csv")

d <- read_csv("250000540.csv")

# during analysis, it was found that the sites located at the following British National Grid points did not have enough data for each species to fit the model (373500, 819500) and (311500, 639500). The following code removes those. 

d <- d |>
  filter(!(x == 373500 & y == 819500) & 
         !(x == 311500 & y == 639500))
```


# Part A

## Occupancy Model

### Data Wrangling

```{r}
grids <- d |> 
  distinct(x, y) |> 
  arrange(x, y)

y_list <- d |>
  select(x, y, species, visit, obs) |>
  pivot_wider(names_from = visit, values_from = obs, names_prefix = "obs_") |>
  arrange(x, y) |>
  split(~ species) |>
  lapply(\(df) df |> select(starts_with("obs")) |> as.matrix())

# view <- data.frame(y_list$A, y_list$B)

site_covs <- grids |>
  left_join(tot_landcover, by = c("x", "y")) |>
  select(Urban, Coniferous, Broadleaf, MountainHeathBog, Freshwater) |>
  as.data.frame()

obs_wide <- d |>
  distinct(x, y, visit, rain, wind) |>
  arrange(x, y, visit) |>
  pivot_wider(names_from = visit, values_from = c(rain, wind))

obs_covs <- list(
  rain = obs_wide |> select(starts_with("rain")) |> as.data.frame(),
  wind = obs_wide |> select(starts_with("wind")) |> as.data.frame()
)
```

### Model Code / Eval

```{r}
umf <- unmarkedFrameOccuMulti(y = y_list, # observations
                         siteCovs = site_covs, # site covariates
                         obsCovs = obs_covs) # observation covariates
```

```{r}
summary(umf)
```

```{r, warning=F}
m1 <- occuMulti(detformulas = c("~1", "~1"), 
                stateformulas = c("~1", "~1", "~1"), 
                data = umf)

m2 <- occuMulti(detformulas = c("~wind + rain", "~wind + rain"), 
                stateformulas = c("~1", "~1", "~1"), 
                data = umf)

m3 <- occuMulti(detformulas = c("~1", "~1"), 
                stateformulas = c("~Urban + Coniferous + Broadleaf + MountainHeathBog + Freshwater",
                                  "~Urban + Coniferous + Broadleaf + MountainHeathBog + Freshwater",
                                  "~1"), 
                data = umf)

m4 <- occuMulti(detformulas = c("~wind + rain", "~wind + rain"), 
                stateformulas = c("~Urban + Coniferous + Broadleaf + MountainHeathBog",
                                  "~Urban + Coniferous + Broadleaf + MountainHeathBog",
                                  "~1"), 
                data = umf)

m5 <- occuMulti(detformulas = c("~wind + rain", "~wind + rain"), 
                stateformulas = c("~Urban + Coniferous + Broadleaf + Freshwater",
                                  "~Urban + Coniferous + Broadleaf + Freshwater",
                                  "~1"), 
                data = umf)

m6 <- occuMulti(detformulas = c("~wind + rain", "~wind + rain"), 
                stateformulas = c("~Urban + Coniferous + MountainHeathBog + Freshwater",
                                  "~Urban + Coniferous + MountainHeathBog + Freshwater",
                                  "~1"), 
                data = umf)

m7 <- occuMulti(detformulas = c("~wind + rain", "~wind + rain"), 
                stateformulas = c("~Urban + Broadleaf + MountainHeathBog + Freshwater",
                                  "~Urban + Broadleaf + MountainHeathBog + Freshwater",
                                  "~1"), 
                data = umf)

m8 <- occuMulti(detformulas = c("~wind + rain", "~wind + rain"), 
                stateformulas = c("~Coniferous + Broadleaf + MountainHeathBog + Freshwater",
                                  "~Coniferous + Broadleaf + MountainHeathBog + Freshwater",
                                  "~1"), 
                data = umf)

m9 <- occuMulti(detformulas = c("~wind + rain", "~wind + rain"), 
                stateformulas = c("~Urban + Coniferous + Broadleaf + MountainHeathBog + Freshwater",
                                  "~Urban + Coniferous + Broadleaf + MountainHeathBog + Freshwater",
                                  "~1"), 
                data = umf) 

mod_list <- fitList(m1 = m1, m2 = m2, m3 = m3, m4 = m4, m5 = m5, 
                    m6 = m6, m7 = m7, m8 = m8, m9 = m9)
```

```{r}
eval_list <- modSel(mod_list)
eval_list
```

This basic variable selection check shows that the strongest model based on AIC is the model where all site and environmental variables are considered.

```{r}
summary(m3)
```

By iteratively removing the least significant predictor until all are significant the alpha = 0.05 level, the final model discounts the rain, urban, and freshwater predictors. 

```{r, warning=F}
final_occ_mod <- occuMulti(detformulas = c("~1", "~1"), 
                stateformulas = c("~ Urban + Broadleaf + MountainHeathBog",
                                  "~ Coniferous + Broadleaf + Freshwater",
                                  "~1"), 
                data = umf)

summary(final_occ_mod)
```

This is the lowest AIC model.

```{r}
# Predict separately for each species using the deduplicated site_covs
preds_a <- predict(final_occ_mod, type = "state", species = "A", newdata = site_covs) |>
  bind_cols(grids) |> 
  mutate(species = "Species A")

preds_b <- predict(final_occ_mod, type = "state", species = "B", newdata = site_covs) |>
  bind_cols(grids) |> 
  mutate(species = "Species B")

# Combine, convert to sf, and project
preds_sf <- bind_rows(preds_a, preds_b) |>
  st_as_sf(coords = c("x", "y"), crs = 27700) |>
  st_transform(crs = 4326)

# Build a single faceted map
occ_map <- ggplot() +
  geom_sf(data = ma, size = 0.05) +
  geom_sf(data = preds_sf, aes(color = Predicted)) +
  scale_color_viridis_b() +
  facet_wrap(~ species) +
  theme_bw() +
  labs(color = "Predicted Occupancy Probability") +
  theme(legend.position = "bottom") 

occ_map

ggsave("occ_map_combined.png", plot = occ_map, path = here::here("A3"),
       width = 10, height = 6, units = "in")
```









